---
title: "KINASE ACTIVITY PROFILING ANALYSIS"
author: "MINA PEYTON"
date: '2022-05-01'
output: html_document
---
# 1. Import Kinase-substrate annotaiton database

To perform Kinase Activity Profiling Analysis, we first downloaded the kinase 
substrate annotation database from PhosphoSitePlus (www.phosphosite.org). 
```{r}
# import kinase substrate annotation database from PhosphoSitePlus 
Kinase_Substrate_Dataset <- read.delim("~/Desktop/KAPA/Kinase_Substrate_Dataset")

head(Kinase_Substrate_Dataset, 10)
```

```{r}
# extract Kin_organism == rat
KS_rat <-subset(Kinase_Substrate_Dataset, Kinase_Substrate_Dataset$KIN_ORGANISM == "rat")

head(KS_rat,10)
```

# 2. Import Aged phosphoproteome dataset 

```{r}
data_sub <- read.csv("Adata_substrates.csv") 
data_sub2 <- read.csv("Ydata_substrates.csv")

```

# 3. Determine specific kinases and substrates identified in phosphoproteome dataset
Data preprocessing to identify kinases and substrates specific to the dataset

```{r}
# extract kinase-substrates from annotation and data files by Gene name 
kinase_subGene <-KS_rat$SUB_GENE 
data_subGene <- data_sub$GeneID 
data_subGene2 <- data_sub2$GeneID 

# which of the kinase-substrates are found in our dataset
subGene <- KS_rat[which(kinase_subGene %in% data_subGene),] #Aged
subGene2 <- KS_rat[which(kinase_subGene %in% data_subGene2),] #Young

head(subGene,5)

```


```{r}
# which kinases do these substrates map to that is unique to our dataset
Dkinases <- unique(subGene$KINASE) # 
Dkinases2 <- unique(subGene2$KINASE) # 
Dkinases
```

# 4. Extract substrate phosphorylation ratios & k phospho-sites per kinase
To assess the activity of each kinase, we extracted the substrates site
phosphorylation ratios of identified kinase substrates based on the 
kinase–substrate annotation profile. 

```{r}
# create for loop to extract the 
# phosphorylation log2ratio of identified kinase substrates 
# based on the kinase–substrate annotation profile & extract k phospho sites 
# identified per kinase

m.results <- c()
k.results <- c()

for (i in seq_along(Dkinases)){
  substrates <- KS_rat[which(KS_rat$KINASE == Dkinases[i]),] 
  subndata <- data_sub[which(data_sub$GeneID %in% substrates$SUB_GENE),]
  k.results[i] <- as.numeric(dim(subndata)[1])
  m.results[i] <- mean(as.numeric(subndata$Log2ratio))}

# Aged
substrates <- KS_rat[which(KS_rat$KINASE == Dkinases),] 
subndata <- data_sub[which(data_sub$GeneID %in% substrates$SUB_GENE),]

for (i in seq_along(Dkinases2)){
  substrates <- KS_rat[which(KS_rat$KINASE == Dkinases2[i]),] 
  subndata <- data_sub2[which(data_sub2$GeneID %in% substrates$SUB_GENE),]
  k.results[i] <- as.numeric(dim(subndata)[1])
  m.results[i] <- mean(as.numeric(subndata$Log2ratio))}
```

# 5. Perform bootstrapping-based random sampling. 
For k target sites of a specific kinase substrate quantified in this study, randomly select k quantifiable phosphorylation sites from dataset, and calculate average log2 ratio of the randomly selected phosphorylation sites. 

Random selection process was repeated 10,000 times. The average log2 ratios from these random selections will form a normal distribution based on the Central Limit Theorem. 

```{r}
# create loop for k target sites of a specific kinase substrate quantified
# to perform bootstrapping-based random sampling

data_log2ratio <- as.numeric(data_sub$Log2ratio)
data_log2ratio <- as.numeric(data_sub2$Log2ratio)
n.samples <- 10000 
bootstrap.results <- c() 
bootstrap.avgs <- list() 

write.csv(data_log2ratio, "Aged_log2ratio.csv")
write.csv(data_log2ratio, "Young_log2ratio.csv")

set.seed(2) # # 2 == A all, 3 == Y all
for (k in k.results){
  for (i in 1:n.samples){
    obs <- sample(data_log2ratio,k, replace=TRUE)
    bootstrap.results[i] <- mean(obs) 
    avgs <- list(bootstrap.results)}
  bootstrap.avgs <- append(bootstrap.avgs,avgs)
}

length(bootstrap.results) # Sanity check: this should contain the mean of 10000 different samples
length(bootstrap.avgs) 
hist(bootstrap.results) 

```

# Kinase Activity Score
Calculate the z-score for the average log2 ratio of the specific kinase target sites

```{r}
### create loop to calculate kinase activity score
bootmean <- c()
bootsd <- c()
K_score <- c() # Aged
K_score2 <-c() # Young

for (i in seq_along(bootstrap.avgs)){
  avg <- mean(bootstrap.avgs[[i]])
  sd <- sd(bootstrap.avgs[[i]])
  bootmean[i] <- avg
  bootsd[i] <- sd
  K_score[i] <- (m.results[i] - bootmean[i])/bootsd[i]
}

barplot(K_score) # Aged

for (i in seq_along(bootstrap.avgs)){
  avg <- mean(bootstrap.avgs[[i]])
  sd <- sd(bootstrap.avgs[[i]])
  bootmean[i] <- avg
  bootsd[i] <- sd
  K_score2[i] <- (m.results[i] - bootmean[i])/bootsd[i]
}

barplot(K_score2) # Young


```
# Calculate the p-values from Z-score == determine which kinases are statistically significant, p < 0.05

```{r}

Kpvalue1 <-c() # aged
Kpvalue2 <- c() #young

# Create a loop to calculate kinase pvalues
# Aged
for (i in K_score){
  pvalues <- (1-pnorm(abs(i)))*2
  Kpvalue1 <-c(Kpvalue1, pvalues)
}

# Young
for (i in K_score2){
  pvalues <- (1-pnorm(abs(i)))*2
  Kpvalue2 <-c(Kpvalue2, pvalues)
}

K1p.adjust <- p.adjust(Kpvalue1, method = "BH")
K2p.adjust <- p.adjust(Kpvalue2, method = "BH")

sigKinases1 <- cbind(Dkinases,K_score,Kpvalue1, K1p.adjust)
sigKinases2 <- cbind(Dkinases2,K_score2,Kpvalue2, K2p.adjust)

signK1 <- sigKinases1[which(sigKinases1[,3] < 0.05),]
signK1 # Aged
write.csv(signK1, "Aged_sigKinases.csv")
```
