---
title: "Kinase Activity Profiling Analysis"
output:
  html_document:
    df_print: paged
  pdf_document: default
  html_notebook: default
---

# 1. Import Kinase-substrate annotaiton database

To perform Kinase Activity Profiling Analysis, we first downloaded the kinase 
substrate annotation database from PhosphoSitePlus (www.phosphosite.org). 
```{r}
# import kinase substrate annotation database from PhosphoSitePlus 
Kinase_Substrate_Dataset <- read.delim("~/Desktop/Choi_phospho/Phospho_analysis/KAPA/Kinase_Substrate_Dataset")

head(Kinase_Substrate_Dataset, 10)
```

```{r}
# extract Kin_organism == rat
KS_rat <-subset(Kinase_Substrate_Dataset, Kinase_Substrate_Dataset$KIN_ORGANISM == "rat")

head(KS_rat,10)
```

```{r}
unique(KS_rat$GENE) # 106 unique kinases in KS annotation file
```

# 2. Import Aged phosphoproteome dataset 

```{r}
data_sub <- read.csv("Adata_substrates.csv") #10465
data_sub2 <- read.csv("Ydata_substrates.csv") #8343

```

# 3. Determine specific kinases and substrates identified in phosphoproteome dataset
Data preprocessing to identify kinases and substrates specific to the dataset

```{r}
# extract kinase-substrates from annotation and data files by Gene name 
kinase_subGene <-KS_rat$SUB_GENE #1386, annotation
data_subGene <- data_sub$GeneID # 10465, Aged
data_subGene2 <- data_sub2$GeneID # 8343, Young

# which of the 1386 kinase-substrates are found in our dataset
subGene <- KS_rat[which(kinase_subGene %in% data_subGene),] #Aged
subGene2 <- KS_rat[which(kinase_subGene %in% data_subGene2),] #Young
#700 in Aged
#686 in Young
head(subGene,5)

```


```{r}
# which kinases do these substrates map to that is unique to our dataset
Dkinases <- unique(subGene$KINASE) # 77 kinases in aged
Dkinases2 <- unique(subGene2$KINASE) # 76 kinases in young

Dkinases
```

# 4. Extract substrate phosphorylation ratios & k phospho-sites per kinase
To assess the activity of each kinase, we extracted the substrates site
phosphorylation ratios of identified kinase substrates based on the 
kinase–substrate annotation profile. 

```{r}
# create for loop to extract the 
# phosphorylation log2ratio of identified kinase substrates 
# based on the kinase–substrate annotation profile & extract k phospho sites 
# identified per kinase

m.results <- c()
k.results <- c()

for (i in seq_along(Dkinases)){
  substrates <- KS_rat[which(KS_rat$KINASE == Dkinases[i]),] 
  subndata <- data_sub[which(data_sub$GeneID %in% substrates$SUB_GENE),]
  k.results[i] <- as.numeric(dim(subndata)[1])
  m.results[i] <- mean(as.numeric(subndata$Log2ratio))}

# Aged
substrates <- KS_rat[which(KS_rat$KINASE == Dkinases),] 
subndata <- data_sub[which(data_sub$GeneID %in% substrates$SUB_GENE),]

for (i in seq_along(Dkinases2)){
  substrates <- KS_rat[which(KS_rat$KINASE == Dkinases2[i]),] 
  subndata <- data_sub2[which(data_sub2$GeneID %in% substrates$SUB_GENE),]
  k.results[i] <- as.numeric(dim(subndata)[1])
  m.results[i] <- mean(as.numeric(subndata$Log2ratio))}
```

# 5. Perform bootstrapping-based random sampling. 
For k target sites of a specific kinase substrate quantified in this study, randomly select k quantifiable phosphorylation sites from dataset, and calculate average log2 ratio of the randomly selected phosphorylation sites. 

Random selection process was repeated 10,000 times. The average log2 ratios from these random selections will form a normal distribution based on the Central Limit Theorem. 

```{r}
# create loop for k target sites of a specific kinase substrate quantified
# to perform bootstrapping-based random sampling

data_log2ratio <- as.numeric(data_sub$Log2ratio)
data_log2ratio <- as.numeric(data_sub2$Log2ratio)
n.samples <- 10000 # Number of bootstrap samples
bootstrap.results <- c() # Create empty vector to hold the results
bootstrap.avgs <- list() # create empty list to hold 10000 avgs per kinase

write.csv(data_log2ratio, "Aged_log2ratio.csv")
write.csv(data_log2ratio, "Young_log2ratio.csv")

set.seed(3) # # 2 == A all, 3 == Y all
for (k in k.results){
  for (i in 1:n.samples){
    obs <- sample(data_log2ratio,k, replace=TRUE)
    bootstrap.results[i] <- mean(obs) # Mean of the bootstrap sample
    avgs <- list(bootstrap.results)}
  bootstrap.avgs <- append(bootstrap.avgs,avgs)
}

length(bootstrap.results) # Sanity check: this should contain the mean of 1000 different samples
length(bootstrap.avgs) # kinase elements x 10000 list length
hist(bootstrap.results) # normal distribution

```

# Kinase Activity Score
Calculate the z-score for the average log2 ratio of the specific kinase target sites

```{r}
### create loop to calculate kinase activity score
bootmean <- c()
bootsd <- c()
K_score <- c() # Aged
K_score2 <-c() # Young

for (i in seq_along(bootstrap.avgs)){
  avg <- mean(bootstrap.avgs[[i]])
  sd <- sd(bootstrap.avgs[[i]])
  bootmean[i] <- avg
  bootsd[i] <- sd
  K_score[i] <- (m.results[i] - bootmean[i])/bootsd[i]
}

barplot(K_score) # Aged

for (i in seq_along(bootstrap.avgs)){
  avg <- mean(bootstrap.avgs[[i]])
  sd <- sd(bootstrap.avgs[[i]])
  bootmean[i] <- avg
  bootsd[i] <- sd
  K_score2[i] <- (m.results[i] - bootmean[i])/bootsd[i]
}

barplot(K_score2) # Young


```
# Calculate the p-values from Z-score == determine which kinases are statistically significant, p < 0.05

```{r}

Kpvalue1 <-c() # aged
Kpvalue2 <- c() #young

# Create a loop to calculate kinase pvalues
# Aged
for (i in K_score){
  pvalues <- (1-pnorm(abs(i)))*2
  Kpvalue1 <-c(Kpvalue1, pvalues)
}

# Young
for (i in K_score2){
  pvalues <- (1-pnorm(abs(i)))*2
  Kpvalue2 <-c(Kpvalue2, pvalues)
}

K1p.adjust <- p.adjust(Kpvalue1, method = "BH")
K2p.adjust <- p.adjust(Kpvalue2, method = "BH")

sigKinases1 <- cbind(Dkinases,K_score,Kpvalue1, K1p.adjust)
sigKinases2 <- cbind(Dkinases2,K_score2,Kpvalue2, K2p.adjust)

signK1 <- sigKinases1[which(sigKinases1[,3] < 0.05),]
signK1 # Aged
write.csv(signK1, "Aged_sigKinases.csv")
```

```{r}
signK2 <- sigKinases2[which(sigKinases2[,3] < 0.05),]
signK2 # Young
write.csv(signK2, "Young_sigKinases.csv")
```

```{r}
sigKinases1[which(sigKinases1[,1] == "PKCD"),] # Aged
```


```{r}

sigKinases2[which(sigKinases2[,1] == "PKCD"),] # Young

```
